The Strong Gravitational Lens Finding Challenge

### 2017-06-19 paper.tex
Update on Overleaf.

Strong gravitational lenses are rare cases where a distant galaxy or quasar is aligned so closely with a foreground galaxy or cluster of galaxies  that the gravitational field of the foreground object creates multiple, highly distorted images of the background object.  These take the form of multiple images of a quasar and Einstein rings surrounding a galaxy and of highly distorted arcs seen through galaxy groups and clusters.  The first strong lens was discovered in 1979 \citep{1979Natur.279..381W} and since then many valuable scientific uses have been found for them.  They have been used to study the mass distribution in galaxies and substructure in dark matter.  They have been used to measure the Hubble constant and other cosmological parameters.
Their magnification has been used as a natural telescope to observe otherwise undetectable objects.   They have put limits on the self interaction of dark matter and on the theory of gravitation.  Through microlensing they have been used to study the structure of quasars.   To access this wealth of information the first step is to find the lenses.  \red{Could use some citations in this paragraph.}

@@

More recently machine learning techniques that have become widely used in the fields computer image processing and artificial intelegence 
have been applied to this problem and others in astronomy, in particular, artificial neural networks (ANNs) and support vector machines (SVM). These algorithms do not pick out particular predefined features, but learn from a training set which features are most important for distinguishing categories of objects.  These have been used widely for such tasks as handwriting and facial recognition.  In astronomy they are beginning to be used for catagorizing galaxies morphologies \citep{2015MNRAS.450.1441D}, photometric redshifts \citep{2017MNRAS.465.1959C,2016PASP..128j4502S,2017NewA...51..169S}, supernova classification \citep{2016ApJS..225...31L} and the lens finding problem \citep{2017arXiv170207675P,2017MNRAS.465.4325O}.

Given the future of this field with large amounts of data coming soon and many new ideas emerging we thought it would be a good time to stage a series of challenges to stimulate new work, determine what can realistically be done in lens finding and get a better idea of the strengths and weaknesses to different methods.  The long term goal is to get a set of algorithms that can handle Euclid, LSST or SKA data sets and produce high purity and high completeness lens samples with well defined efficiency or selection.  We anticipate further challenges in the future were the realism of the data simulations will become progressively more realistic.  Here we have chosen to concentrate on galaxy/small group scale lenses where the background source is a galaxy because we feel that this is were the most progress can be made and the scientific return is the highest, but QSO lenses and cluster/group lenses challenges may follow.

@@

The challenge was in fact two separate challenges that could be entered independently.  One was designed to mimic a single band of a future imaging data set from a satellite survey such as Euclid \citep{***} \red{(the latest description)}.  The other was designed to mimic ground based data with multiple bands, roughly modeled on the Kilo-Degree Survey (KiDS)\footnote{http://kids.strw.leidenuniv.nl/} \citep{2013ExA....35...25D} .    In neither case were the simulated images meant to precisely mock these surveys, but the surveys were used as guides to set noise levels, pixel sizes, sensitivities and other parameters.

@@

Creating the mock images starts with a cosmological nbody simulation.  In this case it was the Millennium simulation \citep{2009MNRAS.398.1150B}.   A catalog of dark matter halos within a light-cone is constructed within the Millennium Observatory project \citep{2013MNRAS.428..778O}.   The challenge sets were based on a 1.6 sq.deg. light cone extending out to redshift 6.
The halo catalog is read into the GLAMER lensing code \citep{2014MNRAS.445.1942M,2014MNRAS.445.1954P} to do all the ray-tracing.  This code projects the halos onto lens planes, in this case 20, and finds the deflection angle on each plane by a hybrid tree method.  In this way the halos have the mass, concentration and clustering properties from the nbody simulation, but within each strong lens the mass resolution is not limited by the original simulation, but follows the analytic mass profile.

@@

 A **** square degree light-cone gives rise to **** such caustics, but these over count the number of lenses since most will have caustics on multiple source planes.  These lenses could be used as is, but we wanted to produce a much larger number with more randomness.   For each caustic we identify the lens plane with the highest convergence and identify all the halos within a three dimensional distance of 0.5 Mpc from the center of the critical curve and on this and its neighboring lens planes.    This collection of halos is then used as the lens and rotated to produce more random lenses.    It  contains all the subhalos and nearby companion halos, but not the large scale structure surrounding it.

For lensed sources we use sources from the Hubble Ultra Deep Field (UDF) that have been decomposed into shapelet functions to remove noise.  This is the same set of images as used 
in  \red{*** Massimo: Can you put a citation in for this?}.   There are 9,350 such sources with redshifts and separate shapelet coefficients in 10 bands. \red{(verify)}

@@

The visible galaxies associated with the lens must also be simulated.
There were not enough large galaxies in the UDF catalog to make enough mock lens galaxies for this purpose.  Instead, for most of the lenses, we used an analytic model for the surface brightness of these galaxies.  The Millennium Observatory provides parameters for the galaxies that inhabit the dark matter halos using the semi-analytic galaxy formation models of \citet{2011MNRAS.413..101G}.  The parameters used here were the total magnitude, the bulge-to-disk ratio, the disk scale height and the bulge effective radius.  The magnitude and bulge-to-disk ratio are a function of the pass band.  Each galaxy is given a random orientation and inclination angle.between 0 and $80\degree$.  The disk is exponential with no vertical height which is why the inclination is limited to $80\degree$.  The bulge is represented by a S\'{e}rsic profile with a axis ratio randomly sampled between 0.5 and 1.  The S\'{e}rsic index, $n_s$, is given by 
\begin{align}
\log( n_s )= 0.4 \log\left[ {\rm max}\left(\frac{B}{T},0.03\right)\right] + 0.1 x 
\end{align}
\red{(reference)} where $\frac{B}{T}$ is the bulge to total flux ratio and $x$ is a uniform random number between -1 and 1.  This approximately reproduces the observed correlation between these quantities.

@@

To represent the mass of the galaxies we make a gridded map of the surface brightness at 3 times the resolution of the final image.  The surface brightness map in converted into a mass map by assuming a uniform mass-to-light ratio of 1.5 solar in the reference band.  This mass map is added to the lens before the ray tracing is done.

@@

\subsection{space-based}
\label{sec:sim-space-based}

For the space-based datasets the pixel size was set to 0.1 arcsec and 
a simple Gaussian PSF was applied with a FWHM of 0.18 arcsec.  
The reference band was SDSS I.  The noise is based on a 1,800 second exposure, a transmission of 0.3, a telescope diameter of 119~cm, a readout noise of 5 electrons per pixel and a sky background of 22.8 magnitudes per sq.arcsec.  
We use a simple Gaussian random realization of the noise uncorrelated between pixels.

@@

\subsection{ground-based}
\label{sec:sim-ground-based}

For the ground-based images four bands (I, G, U and R) where simulated.  The reference band was R.  For the challenge set 85\% images where made with purely simulated images as outlines above and the other 15\% used actual images taken from a preliminary sample of bright galaxies directly from the KiDS survey.  Lensed source images where added to these real images at the same rate as for the mock images,  in this case 50\%.  No attempt was made to match the halo masses to the observed galaxies in these cases.  These real images where added for more realism and so that, by comparing the results for real and mock images, we can evaluate how realistic our simulations are in this context.  There were about 160,000 of these stamps from KiDS.

The KiDS survey provided a representative PSF map in each band that was applied to all mock images.   The pixel size in this case was 0.2~arcsec. 
Weight maps for the KiDS images were also provided.  Some of these had masked regions from remove stars, cosmic rays and bad pixels.  For the mock images the noise was simulated by adding normally distributed numbers with the variance given by the weight maps.  The weight maps where also randomly rotated and flipped.  This resulted in many of the images having large masked regions in them.

@@

\section{the entries}
\label{sec:entries}

@@

The network itself consists of three convolutional layers ($11\times 11\times 32$, $5\times 5 \times 64$ and $3\times 3\times 64$), followed by two fully-connected layers ($256$ and $64$ neurons) and an output softmax layer. The first five layers use the ELU (Exponential Linear Unit) activation function \citep{2015arXiv151107289C}, which in our tests led to significantly faster convergence compared to ReLU and even SoftPlus activation. Dropout regularization \citep{2012arXiv1207.0580H,JMLR:v15:srivastava14a} is applied to both convolutional and fully connected layers, with ``keep'' probabilities $p=2/3$ and $p=1/2$, respectively.

@@

@@

\bibliographystyle{mn2e}
\bibliography{references}

>>>>

Strong gravitational lenses are rare cases where a distant galaxy or quasar is aligned so closely with a foreground galaxy or cluster of galaxies  that the gravitational field of the foreground object creates multiple, highly distorted images of the background object.  These take the form of multiple images of a quasar and Einstein rings surrounding a galaxy and of highly distorted arcs seen through galaxy groups and clusters.  The first strong lens was discovered in 1979 \citep{1979Natur.279..381W} and since then many valuable scientific uses have been found for them.  They have been used to study how dark matter is distributed in galaxies and clusters \citep[e.g.][]{1991ApJ...373..354K,2001ApJ...554.1216C,2002ApJ...568L...5K,2003ApJ...587..143R,2003ApJ...583..606K,2005MNRAS.360.1333W,2005ApJ...623...31D,2009MNRAS.392..945V,2016MNRAS.463.3115T}.  They have been used to measure the Hubble constant and other cosmological parameters \citep[e.g.][]{1964MNRAS.128..307R,1992ARA&A..30..311B,2000ApJ...544...98W,2013ApJ...766...70S,2016A&ARv..24...11T}.
Their magnification has been used as a natural telescope to observe otherwise undetectable objects \citep[e.g][]{2007ApJ...671.1196M,2017MNRAS.464.4823B,2016ApJ...833..264S}.   They have put limits on the self interaction of dark matter and on the theory of gravitation.  Through microlensing they have been used to study the structure of quasars.   To access this wealth of information the first step is to find the lenses.  \red{Could use some citations in this paragraph.}

@@

More recently machine learning techniques that have become widely used in the fields computer image processing and artificial intelligence 
have been applied to this problem and others in astronomy, in particular, artificial neural networks (ANNs) and support vector machines (SVM). These algorithms do not pick out particular predefined features, but learn from a training set which features are most important for distinguishing categories of objects.  These have been used widely for such tasks as handwriting and facial recognition.  In astronomy they are beginning to be used for catagorizing galaxies morphologies \citep{2015MNRAS.450.1441D}, photometric redshifts \citep{2017MNRAS.465.1959C,2016PASP..128j4502S,2017NewA...51..169S}, supernova classification \citep{2016ApJS..225...31L} and the lens finding problem \citep{2017arXiv170207675P,2017MNRAS.465.4325O}.

Given the future of this field with large amounts of data coming soon and many new ideas emerging, it is timely to stage a series of challenges to stimulate new work, determine what can realistically be done in lens finding and get a better idea of the strengths and weaknesses to different methods.  The long term goal is to get a set of algorithms that can handle Euclid, LSST or SKA data sets and produce high purity and high completeness lens samples with well defined efficiency or selection.  We anticipate further challenges in the future were the realism of the data simulations will become progressively larger.  Here we have chosen to concentrate on galaxy/small group scale lenses where the background source is a galaxy because we feel that this is where the most progress can be made and the scientific return is the highest, but QSO lenses and cluster/group lenses challenges may follow.

@@

The challenge was in fact two separate challenges that could be entered independently.  One was designed to mimic a single band of a future imaging data set from a satellite survey such as Euclid \citep{***} \red{(the latest description)}.  The other was designed to mimic ground based data with multiple bands, roughly modeled on the Kilo-Degree Survey (KiDS)\footnote{http://kids.strw.leidenuniv.nl/} \citep{2013ExA....35...25D}. In neither case were the simulated images meant to precisely mock these surveys, but the surveys were used as guides to set noise levels, pixel sizes, sensitivities, and other parameters.

@@

Creating the mock images starts with a cosmological N-body simulation.  In this case it was the Millennium simulation \citep{2009MNRAS.398.1150B}.   A catalog of dark matter halos within a light-cone is constructed within the Millennium Observatory project \citep{2013MNRAS.428..778O}.   The challenge sets were based on a 1.6 sq.deg. light cone extending out to redshift $z=6$.
The halo catalog is read into the GLAMER lensing code \citep{2014MNRAS.445.1942M,2014MNRAS.445.1954P} to do all the ray-tracing.  This code projects the halos onto lens planes, in this case 20, and finds the deflection angle on each plane by a hybrid tree method.  In this way the halos have the mass, concentration and clustering properties from the N-body simulation, but within each strong lens the mass resolution is not limited by the original simulation, but follows the analytic mass profile. {\color{red} MM: I am a bit confused: at this stage, how are the halos modeled? N-body or analytic? What is the shape of the halos?}

@@

 A **** square degree light-cone gives rise to **** such caustics, but these over count the number of lenses since most will have caustics on multiple source planes.  These lenses could be used as is, but we wanted to produce a much larger number with more randomness.   For each caustic we identify the lens plane with the highest convergence and identify all the halos within a three dimensional distance of 0.5 Mpc from the center of the critical curve and on this and its neighboring lens planes.    This collection of halos is then used as the lens and rotated to produce more random lenses.    It  contains all the sub-halos and nearby companion halos, but not the large scale structure surrounding it.

For lensed sources we use sources from the Hubble Ultra Deep Field (UDF) that have been decomposed into shapelet functions to remove noise.  This is the same set of images as used 
in  \citep{2008A&A...482..403M,2010A&A...514A..93M,2016arXiv160604548M}.   There are 9,350 such sources with redshifts and separate shapelet coefficients in 4 bands.

@@

The visible galaxies associated with the lens must also be simulated. There  too few bright galaxies in the  in the UDF catalog to make enough mock lens galaxies for this purpose.  Instead, for most of the lenses, we used an analytic model for the surface brightness of these galaxies.  The Millennium Observatory provides parameters for the galaxies that inhabit the dark matter halos using the semi-analytic galaxy formation models of \citet{2011MNRAS.413..101G}.  The parameters used here were the total magnitude, the bulge-to-disk ratio, the disk scale height and the bulge effective radius.  The magnitude and bulge-to-disk ratio are a function of the pass band.  Each galaxy is given a random orientation and inclination angle.between 0 and $80\degree$.  The disk is exponential with no vertical height which is why the inclination is limited to $80\degree$.  The bulge is represented by a S\'{e}rsic profile with a axis ratio randomly sampled between 0.5 and 1.  The S\'{e}rsic index, $n_s$, is given by 
\begin{align}
\log( n_s )= 0.4 \log\left[ {\rm max}\left(\frac{B}{T},0.03\right)\right] + 0.1 x 
\end{align}
\red{(reference)} where $\frac{B}{T}$ is the bulge to total flux ratio and $x$ is a uniform random number between -1 and 1.  This approximately reproduces the observed correlation between these quantities.

@@

To represent the mass of the galaxies we make a gridded map of the surface brightness at 3 times the resolution of the final image.  The surface brightness map in converted into a mass map by assuming a uniform mass-to-light ratio of 1.5 solar in the reference band.  This mass map is added to the lens before the ray tracing is done. {\color{red} MM: Again, I am confused with how the lenses are modeled in this stage of the simulation and during the search for caustics described in the beginning of this section.}

@@

\subsection{Space-based}
\label{sec:sim-space-based}

For the space-based datasets the pixel size was set to 0.1 arcsec and 
a simple Gaussian PSF was applied with a FWHM of 0.18 arcsec.  
The reference band was SDSS $i$.  The noise is based on a 1,800 second exposure, a transmission of 0.3, a telescope diameter of 119~cm, a readout noise of 5 electrons per pixel and a sky background of 22.8 magnitudes per sq.arcsec.  
We use a simple Gaussian random realization of the noise uncorrelated between pixels.

@@

\subsection{Ground-based}
\label{sec:sim-ground-based}

For the ground-based images four bands ($u,g,r$, and $i$) where simulated {\color{red} (these are SDSS filters too, right?)}.  The reference band was $r$.  For the challenge set 85\% images where made with purely simulated images as outlines above and the other 15\% used actual images taken from a preliminary sample of bright galaxies directly from the KiDS survey.  Lensed source images where added to these real images at the same rate as for the mock images,  in this case 50\%.  No attempt was made to match the halo masses to the observed galaxies in these cases.  These real images where added for more realism and so that, by comparing the results for real and mock images, we can evaluate how realistic our simulations are in this context.  There were about 160,000 of these stamps from KiDS.

The KiDS survey provided a representative PSF map in each band that was applied to all mock images.   The pixel size in this case was 0.2~arcsec. 
Weight maps for the KiDS images were also provided.  Some of these had masked regions from removed stars, cosmic rays, and bad pixels.  For the mock images the noise was simulated by adding normally distributed numbers with the variance given by the weight maps.  The weight maps where also randomly rotated and flipped.  This resulted in many of the images having large masked regions in them.

@@

\section{The Entries}
\label{sec:entries}

@@

{\color{red} (MM: I think that the co-authors should provide a more detailed description of the algorithms. For example, they should quickly explain what is a CNN and what are its general components (i.e. what kind of layers, activation functions, etc.)}

The network itself consists of three convolutional layers ($11\times 11\times 32$, $5\times 5 \times 64$ and $3\times 3\times 64$), followed by two fully-connected layers ($256$ and $64$ neurons) and an output softmax layer {\color{red} MM: the numbers make little sense without a description of what the mean -- see my previous comment}. The first five layers use the ELU (Exponential Linear Unit) activation function \citep{2015arXiv151107289C}, which in our tests led to significantly faster convergence compared to ReLU and even SoftPlus activation. Dropout regularization \citep{2012arXiv1207.0580H,JMLR:v15:srivastava14a} is applied to both convolutional and fully connected layers, with ``keep'' probabilities $p=2/3$ and $p=1/2$, respectively.

@@

{\red MM: May be we could separate the methods into three categories: machine-learning, arc-finders, visual inspection and group them accordingly}

@@

\bibliographystyle{mnras}
\bibliography{references}
